Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-11-16 mnras.tex
Completed and alphabetized symbol table.

\input macros.tex
\voffset=-2cm

@@

An important case was considered by \citet{Stroemgren39} for an
ultraviolet (UV) radiation source photoionizing a static uniform
neutral medium.  When recombinations balance photo-ionizations, the
radius of a so-called \hii region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_H^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics of
the ISM.  A decade later, the first radiation hydrodynamical numerical
models of \hii regions in spherical symmetry and plane-parallel
ionization fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
fronts are subsonic.  Furthermore ``weak'' and ``strong'' R-type
fronts move supersonically and subsonically with respect to the
ionized gas, respectively.  The same terminology conversely applies to
D-type fronts.  ``Critical'' fronts are defined as moving exactly at
the sound speed.  These works established the evolutionary track of an
expanding \hii illuminated by a massive star in a uniform medium:
%
\begin{enumerate}
\item Weak R-type:  When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type:  As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type:  The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase:  After the shock forms, the \hii region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

@@

\section{Treatment of Radiative Transfer}

@@

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following section, we
assume the rays are monochromatic.  For convenience, we express the
integration in terms of optical depth $\tau = \int \kappa(r,t) \; dr$,
and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  We use the
cell-centered density in our calculations.  Using trilinearly
interpolated densities \citep[see][]{Mellema06} did not produce
improved results.  Equation (\ref{eqn:rtray}) has a simple exponential
analytic solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates can be used to update the ionization state and
energy of the cells.  Considering a system with only hydrogen
photo-ionizations and radiative recombinations, these changes are very
straightforward and is useful for illustrative purposes.  The change
in neutral hydrogen is
%
\begin{equation}
\frac{dn_{\rm H}}{dt} = \alpha_B n_e n_p - C_{\rm H} n_e n_{\rm
  H} - k_{\rm ph},
\end{equation}
where $\alpha_B = 2.59 \times 10^{-13} \mathrm{cm}^3 \;
\mathrm{s}^{-1}$ is the recombination coefficient at 10$^4$ K in the
Case B on-the-spot approximation in which all recombinations are
locally reabsorbed, \citep{Spitzer78}, and $C_{\rm H}$ is the
collisional ionization rate.  However, for more accurate solutions in
calculations that consider several chemical species, the
photo-ionization rates are terms in the relevant chemical networks
\citep[e.g.][]{Abel97}.

@@

For a ray tracing method to accurately, i.e. without non-spherical
artifacts, compute the radiation field, the computational grid must be
well-sampled by the rays.  The main source of potential artifacts is
the geometrical difference between the cell and the HEALPix pixel
relevant in the angular integration of the intensity over the cell.
In this section, we devise a correction scheme to account for these
differences.  Consider the solid angle $\Omega_{\rm ray}$ and photon
flux $P$ associated with a single ray, and assume the flux is constant
across $\Omega_{\rm ray}$.  There exists a discrepancy between the
geometry cell face and HEALPix pixel when the pixel does not cover the
entire cell face, which is illustrated in Figure \ref{fig:covering}.
This mismatch causes non-spherical artifacts and is most apparent in
the optically thin case, where the area of the pixel is dominant over
$(1 - e^\tau)$ when calculating $k_{\rm ph}$.  One can avoid these
artifacts by increasing the sampling $\Phi_c$ to high values ($>10$)
but we have formulated a simple geometric correction to the
calculation of the radiation field.  This correction is not unique to
the HEALPix formalism but can be applied to any type of pixelization.

The contribution to $k_{\rm ph}$ and $\Gamma_{\rm ph}$ must be
corrected by the covering factor $f_c$ of the pixel with respect to
the cell.  When the pixel is fully contained within the cell face,
$f_c \equiv 1$.  Because the geometry of the pixel can be complex with
curved edges, we approximate $f_c$ by assuming the pixel is square.
The covering factor is thus related to the width of a pixel, $L_{\rm
  pix} = R_0 \, \theta_{\rm pix}$, and the distance from the ray
segment midpoint to the closest cell boundary $D_{\rm c,i}$, which is
depicted in Figure \ref{fig:covering}.  To estimate $f_c$, we first
find the distance $d_{\rm center,i}$ from the midpoint of the ray
segment to the cell center $x_{\rm 0,i}$ in orthogonal directions,
%
\begin{equation}
  \label{eqn:midpoint}
  D_{\rm c,i} = \left| R_{\rm 0,i} + \hat{n}_i\,\frac{dr}{2} - x_{\rm
      0,i} \right|,
\end{equation}
where $R_{\rm 0,i}$ is the distance travelled by the ray in each
orthogonal direction.  The distance to the closest cell boundary is
$D_{\rm edge} = dx/2 - \max_{i=1 \rightarrow 3} (D_{\rm c,i})$.  Thus
the covering factor is related to the square of the ratio between the
$L_{\rm pix}$ and $D_{\rm edge}$ by
%
\begin{equation}
  \label{eqn:fc}
  f_c = \left( \frac{1}{2} + \frac{D_{\rm edge}}{L_{\rm pix}} \right)^2
\end{equation}
One half of the pixel is always contained within the cell, which
results in the factor of 1/2.  Finally we multiply $k_{\rm ph}$ and
$\Gamma_{\rm ph}$ by $f_c$ but leave the absorbed radiation $dP$
untouched because this would underestimate the attenuation of the
incoming radiation.  Using $f_c$ calculated like above, the method is
no longer photon conserving.  In our implementation, we felt that the
spherical symmetry obtained outweighed the loss of photon
conservation.  However we show that there are no perceptible
deviations from photon conservation in \S\ref{sec:test1} and
\S\ref{sec:dx_dep}.

@@

High energy photons can also cause Compton heating by scattering off
free electrons.  During a scattering, a photon loses $\Delta E(T_e) =
4kT_e \times (E_{ph} / m_e c^2)$ of energy, where $T_e$ is the
electron temperature.  For the case of monochromatic energy groups, we
model this process by considering that the photons are absorbed by a
factor of 
\begin{equation}
  \frac{dP_C}{P} = (1 - e^{-\tau_e}) \frac{\Delta(T_e)}{E_{ph}},
\end{equation}
which is the equivalent of the photon energy decreasing.  Here $\tau_e
= n_e \sigma_{\rm KN} dl$ is the optical depth to Compton scattering,
and $\sigma_{\rm KN}$ is the non-relativistic Klein-Nishina cross
section \citep{Rybicki}.  The Compton heating rate is thus
\begin{equation}
  \Gamma_{\rm ph,C} = \frac{dP_C}{n_e \; V_{\rm cell} \; dt_P},
\end{equation}
which has been used in \citet{Kim11}.

@@

Another relevant process is radiation pressure, where the absorption
of radiation transfers momentum from photons to the absorbing medium.
This is easily computed by considering the momentum
\begin{equation}
  d\mathbf{p}_\gamma = \frac{dP \; E_{\rm ph}}{c} \; \hat{r}
\end{equation}
of the absorbed radiation from the incoming ray, where $\hat{r}$ is
the normal direction of the ray.  We do not include radiation pressure
on dust currently.  The resulting acceleration of the gas because of
radiation pressure is
\begin{equation}
  d\mathbf{a} = \frac{d\mathbf{p}_\gamma} {dt_P\; \rho \; V_{\rm cell}},
\end{equation}
where $\rho$ is the gas density inside the cell.  This acceleration is
then added to the other forces, e.g. gravity, in the calculation in an
operator split fashion.

\section{Numerical Implementation in Enzo}

@@

\begin{figure*}
  \epsscale{2}
  \plotone{fig/RTcoupling.eps}
  \epsscale{1}
  \caption{\label{fig:MainAlgorithm} Flow chart for the overall
    algorithm of the radiative transfer module in \enzo~that
    illustrates (1) the creation of photon packages, (2) ray tracing,
    (3) the transport of photon packages between AMR grids, and (4)
    coupling with the hydrodynamics.  The ray tracing algorithm, which
    is contained in the ``Trace Rays'' is detailed in Figure
    \ref{fig:RTalgorithm}.}
\end{figure*}

@@

We illustrate the underlying algorithm of the ray tracing module in
\enzo~in Figure \ref{fig:MainAlgorithm} and the ray tracing algorithm
is shown in Figure \ref{fig:RTalgorithm}.  The module is only called
when advancing the finest AMR level.  We describe its steps below.

@@

We search for the cell with the smallest $dt_P$ based on Equation
\ref{eqn:dtelec}.  In principle, one could use this value without
modifications as the timestep, but there is considerable noise both
spatially and temporally.  In order to stabilize this technique, we
first spatially smooth the values of $dt_{\rm P,cell}$ with a Gaussian
filter over a $3^3$ cube.  Because we only consider the cells within
the ionization front, we set $dt_{\rm P,cell}$ to the hydrodynamical
timestep outside the front during the smoothing.  After we have
smoothed $dt_{\rm P,cell}$, we select the minimum value as $dt_P$.
Significant noise in $dt_P$ can exist between time steps as well.
Because the solution can become inaccurate if the timestep is allowed
to be too large, we restrict the next timestep to be less than twice
the previous timestep,
%
\begin{equation}
  \label{eqn:dtchange}
  dt_{\rm P,1} = \min(2 dt_{\rm P,0},\; dt_{\rm P,1}).
\end{equation}
Figure \ref{fig:dtsmooth} shows the smooth evolution of $dt_{\rm P}$
in a growing Str\"{o}mgren sphere when compared to the values of
$\min(dt_{\rm P,cell})$.

@@

For maximum overlap of communication and computation, which enables
scaling to large numbers of processors, we must employ
``non-blocking'' MPI communication, where each processor does not wait
for synchronization with other processors.  We use this technique for
the sending and receiving of rays.  Here we desire to minimize the
idle time of each processor when it is waiting to receive data.  In
the loop shown in Figure \ref{fig:MainAlgorithm} with the conditional
that checks whether we have traced all of the rays, we aggressively
transport rays that are local on the processor, and process any MPI
receive calls as they arrive, not waiting for their completion in
order to continue to the next iteration.  We describe the steps in
this algorithm next.

@@

In Figure \ref{fig:test1_ifront}, we show (a) the evolution of the
neutral and ionization fraction as a function of radius at $t = $ 10,
30, 100, and 500 Myr, and (b) the growth of the ionization front
radius as a function of time and its ratio with the analytical
Str\"{o}mgren radius (Eq. \ref{eqn:rstr}).  The ionization front has a
width of $\sim0.7$ kpc, which is in agreement with the inherent
thickness of $\sim18 \lambda_{\rm mfp} = 0.74$ kpc, given a 13.6 eV
mono-chromatic spectrum.  There are small kinks in the neutral
fraction at 1.5 and 3 kpc that corresponds to artifacts created by the
photon package splitting at these radii.  However these do not affect
the overall solution.  One difference between our results and the
codes presented in RT06 is the increasing neutral fraction outside of
the \hii.  This occurs because the initial ionized fraction and
temperature is set to $1.3 \times 10^{-3}$ and 8000 K, which are not
the equilibrium values.  Over the 500 Myr in the calculation, the
neutral fraction increases to 0.2, which is close to its equilibrium
value.  In the right panel of Figure \ref{fig:test1_ifront}, the
ionization front radius exceeds $R_s$ by a few percent for most of the
calculation.  This difference happens because the analytical solution
(Eq. \ref{eqn:rstr}) assumes the \hii region has a constant
ionized fraction.  The evolution of the ionized fraction as a function
of radius can be analytically calculated \citep[e.g.][]{Osterbrock89,
  Petkova09}, causing the ionization front radius to be slightly
larger, increasing from 0 to 3\% in the interval 80--350 Myr.  Our
results are in excellent agreement with this more accurate analytical
solution.  In Figure \ref{fig:test1_HI}, we show a slice of the
neutral fraction through the origin.  Other than the ray splitting
artifacts that generate the plateaus at 1.5 and 3 kpc, one sees
spherical symmetry without any noise in our solution.

@@

In Figure \ref{fig:test2_1a}, we show the radially averaged neutral
and ionized fraction at $t = $ 10, 30, 100, and 500 Myr, and the total
neutral fraction of the domain in Figure \ref{fig:test2_1b}.  Compared
with Test 1, the ionization front is thicker, as expected with the
harder spectrum.  The total neutral fraction decreases to 0.67 over
4$t_{\rm rec} = 500$ Myr, which is in agreement with the analytical
expectation and other codes in RT06.  In Figure \ref{fig:test2_2}, we
show slices of neutral fraction and temperature through the origin at
$t = $ 10 and 100 Myr.  Here one sees the spherically symmetric \hii
regions and a smooth temperature transition to the neutral ambient
medium.  In Figure \ref{fig:test2_3}, we show the ratio of the
ionization front radius $r_{\rm IF}$ in our simulation and $R_s$.
Before $1.5t_{\rm rec}$, $r_{\rm IF}$ lags behind $R_s$, initially by
10\% and then increases to $R_s$; however afterwards, this ratio
asymptotes to a solution that is 4\% greater than $R_s$.  This
behavior is approximately the median result in RT06, where this ratio
varies between 1 and 1.1, and the early evolution of $r_{\rm IF}$ is
under-predicted by almost all of the codes.  If we use one energy
group with the mean energy (29.6 eV) of a $T=10^5$ K blackbody, we
find that $r_{\rm IF}/R_s$ = 1.08, which is representative of the
codes in the upper range of RT06.

@@

The growth of the ionization front radius is shown in Figure
\ref{fig:test5_1}, using both $T = 10^4$ K and $x_e = 0.5$ as
ionization front definitions, compared to the analytical solution
(Eq. \ref{eqn:test5_final}).  We define this alternative measure
because the ionization front becomes broad as the D-type front creates
a shock.  Densities in this shock, as seen in Figure
\ref{fig:test5_2}, are high enough for the gas to recombine but not
radiatively cool.  Before $2t_{\rm rec} \approx 250$ Myr, the
temperature cutoff overestimates $r_s$ by over 10\%; however at later
times, it provides a good match to the $t^{4/7}$ growth at late times.
With the $x_e = 0.5$ criterion for the ionization front, the radius is
always underestimated by $\sim20\%$.  This behavior was also seen in
RT09.

>>>>

\input macros.tex
\voffset=-1.4cm

@@

An important case was considered by \citet{Stroemgren39} for an
ultraviolet (UV) radiation source photoionizing a static uniform
neutral medium.  When recombinations balance photo-ionizations, the
radius of a so-called \hii region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_{\rm
        H}^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics of
the ISM.  A decade later, the first radiation hydrodynamical numerical
models of \hii regions in spherical symmetry and plane-parallel
ionization fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
fronts are subsonic.  Furthermore ``weak'' and ``strong'' R-type
fronts move supersonically and subsonically with respect to the
ionized gas, respectively.  The same terminology conversely applies to
D-type fronts.  ``Critical'' fronts are defined as moving exactly at
the sound speed.  These works established the evolutionary track of an
expanding \hii illuminated by a massive star in a uniform medium:
%
\begin{enumerate}
\item Weak R-type:  When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type:  As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type:  The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase:  After the shock forms, the \hii region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

@@

\section{Treatment of Radiative Transfer}
\label{sec:rt}

@@

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following section, we
assume the rays are monochromatic.  For convenience, we express the
integration in terms of optical depth $\tau = \int \kappa(r,t) \; dr$,
and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  We use the
cell-centered density in our calculations.  Using trilinearly
interpolated densities \citep[see][]{Mellema06} did not produce
improved results.  Equation (\ref{eqn:rtray}) has a simple exponential
analytic solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates can be used to update the ionization state and
energy of the cells.  Considering a system with only hydrogen
photo-ionizations and radiative recombinations, these changes are very
straightforward and is useful for illustrative purposes.  The change
in neutral hydrogen is
%
\begin{equation}
\label{eqn:dndt}
\frac{dn_{\rm H}}{dt} = \alpha_B n_e n_p - C_{\rm H} n_e n_{\rm
  H} - k_{\rm ph},
\end{equation}
where $\alpha_B = 2.59 \times 10^{-13} \mathrm{cm}^3 \;
\mathrm{s}^{-1}$ is the recombination coefficient at 10$^4$ K in the
Case B on-the-spot approximation in which all recombinations are
locally reabsorbed, \citep{Spitzer78}, and $C_{\rm H}$ is the
collisional ionization rate.  However, for more accurate solutions in
calculations that consider several chemical species, the
photo-ionization rates are terms in the relevant chemical networks
\citep[e.g.][]{Abel97}.

@@

For a ray tracing method to accurately, i.e. without non-spherical
artifacts, compute the radiation field, the computational grid must be
well-sampled by the rays.  The main source of potential artifacts is
the geometrical difference between the cell and the HEALPix pixel
relevant in the angular integration of the intensity over the cell.
In this section, we devise a correction scheme to account for these
differences.  Consider the solid angle $\Omega_{\rm ray}$ and photon
flux $P$ associated with a single ray, and assume the flux is constant
across $\Omega_{\rm ray}$.  There exists a discrepancy between the
geometry cell face and HEALPix pixel when the pixel does not cover the
entire cell face, which is illustrated in Fig. \ref{fig:covering}.
This mismatch causes non-spherical artifacts and is most apparent in
the optically thin case, where the area of the pixel is dominant over
$(1 - e^\tau)$ when calculating $k_{\rm ph}$.  One can avoid these
artifacts by increasing the sampling $\Phi_c$ to high values ($>10$)
but we have formulated a simple geometric correction to the
calculation of the radiation field.  This correction is not unique to
the HEALPix formalism but can be applied to any type of pixelization.

The contribution to $k_{\rm ph}$ and $\Gamma_{\rm ph}$ must be
corrected by the covering factor $f_c$ of the pixel with respect to
the cell.  When the pixel is fully contained within the cell face,
$f_c \equiv 1$.  Because the geometry of the pixel can be complex with
curved edges, we approximate $f_c$ by assuming the pixel is square.
The covering factor is thus related to the width of a pixel, $L_{\rm
  pix} = R_0 \, \theta_{\rm pix}$, and the distance from the ray
segment midpoint to the closest cell boundary $D_{\rm c,i}$, which is
depicted in Fig. \ref{fig:covering}.  To estimate $f_c$, we first
find the distance $d_{\rm center,i}$ from the midpoint of the ray
segment to the cell center $x_{\rm 0,i}$ in orthogonal directions,
%
\begin{equation}
  \label{eqn:midpoint}
  D_{\rm c,i} = \left| R_{\rm 0,i} + \hat{n}_i\,\frac{dr}{2} - x_{\rm
      0,i} \right|,
\end{equation}
where $R_{\rm 0,i}$ is the distance travelled by the ray in each
orthogonal direction.  The distance to the closest cell boundary is
$D_{\rm edge} = dx/2 - \max_{i=1 \rightarrow 3} (D_{\rm c,i})$.  Thus
the covering factor is related to the square of the ratio between the
$L_{\rm pix}$ and $D_{\rm edge}$ by
%
\begin{equation}
  \label{eqn:fc}
  f_c = \left( \frac{1}{2} + \frac{D_{\rm edge}}{L_{\rm pix}} \right)^2
\end{equation}
One half of the pixel is always contained within the cell, which
results in the factor of 1/2.  Finally we multiply $k_{\rm ph}$ and
$\Gamma_{\rm ph}$ by $f_c$ but leave the absorbed radiation $dP$
untouched because this would underestimate the attenuation of the
incoming radiation.  Using $f_c$ calculated like above, the method is
no longer photon conserving.  In our implementation, we felt that the
spherical symmetry obtained outweighed the loss of photon
conservation.  However we show that there are no perceptible
deviations from photon conservation in \S\ref{sec:test1} and
\S\ref{sec:dx_dep}.

@@

High energy photons can also cause Compton heating by scattering off
free electrons.  During a scattering, a photon loses $\Delta E(T_e) =
4kT_e \times (E_{ph} / m_e c^2)$ of energy, where $T_e$ is the
electron temperature.  For the case of monochromatic energy groups, we
model this process by considering that the photons are absorbed by a
factor of 
\begin{equation}
  \label{eqn:compton}
  \frac{dP_C}{P} = (1 - e^{-\tau_e}) \frac{\Delta(T_e)}{E_{ph}},
\end{equation}
which is the equivalent of the photon energy decreasing.  Here $\tau_e
= n_e \sigma_{\rm KN} dl$ is the optical depth to Compton scattering,
and $\sigma_{\rm KN}$ is the non-relativistic Klein-Nishina cross
section \citep{Rybicki}.  The Compton heating rate is thus
\begin{equation}
  \Gamma_{\rm ph,C} = \frac{dP_C}{n_e \; V_{\rm cell} \; dt_P},
\end{equation}
which has been used in \citet{Kim11}.

@@

Another relevant process is radiation pressure, where the absorption
of radiation transfers momentum from photons to the absorbing medium.
This is easily computed by considering the momentum
\begin{equation}
  \label{eqn:radpres}
  d\mathbf{p}_\gamma = \frac{dP \; E_{\rm ph}}{c} \; \hat{r}
\end{equation}
of the absorbed radiation from the incoming ray, where $\hat{r}$ is
the normal direction of the ray.  We do not include radiation pressure
on dust currently.  The resulting acceleration of the gas because of
radiation pressure is
\begin{equation}
  d\mathbf{a} = \frac{d\mathbf{p}_\gamma} {dt_P\; \rho \; V_{\rm cell}},
\end{equation}
where $\rho$ is the gas density inside the cell.  This acceleration is
then added to the other forces, e.g. gravity, in the calculation in an
operator split fashion.

\section{Numerical Implementation in Enzo}
\label{sec:numerical}

@@

\begin{figure*}
  \epsscale{2}
  \plotone{fig/RTcoupling.eps}
  \epsscale{1}
  \caption{\label{fig:MainAlgorithm} Flow chart for the overall
    algorithm of the radiative transfer module in \enzo~that
    illustrates (1) the creation of photon packages, (2) ray tracing,
    (3) the transport of photon packages between AMR grids, and (4)
    coupling with the hydrodynamics.  The ray tracing algorithm, which
    is contained in the ``Trace Rays'' is detailed in Fig.
    \ref{fig:RTalgorithm}.}
\end{figure*}

@@

We illustrate the underlying algorithm of the ray tracing module in
\enzo~in Fig. \ref{fig:MainAlgorithm} and the ray tracing algorithm
is shown in Fig. \ref{fig:RTalgorithm}.  The module is only called
when advancing the finest AMR level.  We describe its steps below.

@@

We search for the cell with the smallest $dt_P$ based on Equation
\ref{eqn:dtelec}.  In principle, one could use this value without
modifications as the timestep, but there is considerable noise both
spatially and temporally.  In order to stabilize this technique, we
first spatially smooth the values of $dt_{\rm P,cell}$ with a Gaussian
filter over a $3^3$ cube.  Because we only consider the cells within
the ionization front, we set $dt_{\rm P,cell}$ to the hydrodynamical
timestep outside the front during the smoothing.  After we have
smoothed $dt_{\rm P,cell}$, we select the minimum value as $dt_P$.
Significant noise in $dt_P$ can exist between time steps as well.
Because the solution can become inaccurate if the timestep is allowed
to be too large, we restrict the next timestep to be less than twice
the previous timestep,
%
\begin{equation}
  \label{eqn:dtchange}
  dt_{\rm P,1} = \min(2 dt_{\rm P,0},\; dt_{\rm P,1}).
\end{equation}
Fig. \ref{fig:dtsmooth} shows the smooth evolution of $dt_{\rm P}$
in a growing Str\"{o}mgren sphere when compared to the values of
$\min(dt_{\rm P,cell})$.

@@

For maximum overlap of communication and computation, which enables
scaling to large numbers of processors, we must employ
``non-blocking'' MPI communication, where each processor does not wait
for synchronization with other processors.  We use this technique for
the sending and receiving of rays.  Here we desire to minimize the
idle time of each processor when it is waiting to receive data.  In
the loop shown in Fig. \ref{fig:MainAlgorithm} with the conditional
that checks whether we have traced all of the rays, we aggressively
transport rays that are local on the processor, and process any MPI
receive calls as they arrive, not waiting for their completion in
order to continue to the next iteration.  We describe the steps in
this algorithm next.

@@

In Fig. \ref{fig:test1_ifront}, we show (a) the evolution of the
neutral and ionization fraction as a function of radius at $t = $ 10,
30, 100, and 500 Myr, and (b) the growth of the ionization front
radius as a function of time and its ratio with the analytical
Str\"{o}mgren radius (Eq. \ref{eqn:rstr}).  The ionization front has a
width of $\sim0.7$ kpc, which is in agreement with the inherent
thickness of $\sim18 \lambda_{\rm mfp} = 0.74$ kpc, given a 13.6 eV
mono-chromatic spectrum.  There are small kinks in the neutral
fraction at 1.5 and 3 kpc that corresponds to artifacts created by the
photon package splitting at these radii.  However these do not affect
the overall solution.  One difference between our results and the
codes presented in RT06 is the increasing neutral fraction outside of
the \hii.  This occurs because the initial ionized fraction and
temperature is set to $1.3 \times 10^{-3}$ and 8000 K, which are not
the equilibrium values.  Over the 500 Myr in the calculation, the
neutral fraction increases to 0.2, which is close to its equilibrium
value.  In the right panel of Fig. \ref{fig:test1_ifront}, the
ionization front radius exceeds $R_s$ by a few percent for most of the
calculation.  This difference happens because the analytical solution
(Eq. \ref{eqn:rstr}) assumes the \hii region has a constant
ionized fraction.  The evolution of the ionized fraction as a function
of radius can be analytically calculated \citep[e.g.][]{Osterbrock89,
  Petkova09}, causing the ionization front radius to be slightly
larger, increasing from 0 to 3\% in the interval 80--350 Myr.  Our
results are in excellent agreement with this more accurate analytical
solution.  In Fig. \ref{fig:test1_HI}, we show a slice of the
neutral fraction through the origin.  Other than the ray splitting
artifacts that generate the plateaus at 1.5 and 3 kpc, one sees
spherical symmetry without any noise in our solution.

@@

In Fig. \ref{fig:test2_1a}, we show the radially averaged neutral
and ionized fraction at $t = $ 10, 30, 100, and 500 Myr, and the total
neutral fraction of the domain in Fig. \ref{fig:test2_1b}.  Compared
with Test 1, the ionization front is thicker, as expected with the
harder spectrum.  The total neutral fraction decreases to 0.67 over
4$t_{\rm rec} = 500$ Myr, which is in agreement with the analytical
expectation and other codes in RT06.  In Fig. \ref{fig:test2_2}, we
show slices of neutral fraction and temperature through the origin at
$t = $ 10 and 100 Myr.  Here one sees the spherically symmetric \hii
regions and a smooth temperature transition to the neutral ambient
medium.  In Fig. \ref{fig:test2_3}, we show the ratio of the
ionization front radius $r_{\rm IF}$ in our simulation and $R_s$.
Before $1.5t_{\rm rec}$, $r_{\rm IF}$ lags behind $R_s$, initially by
10\% and then increases to $R_s$; however afterwards, this ratio
asymptotes to a solution that is 4\% greater than $R_s$.  This
behavior is approximately the median result in RT06, where this ratio
varies between 1 and 1.1, and the early evolution of $r_{\rm IF}$ is
under-predicted by almost all of the codes.  If we use one energy
group with the mean energy (29.6 eV) of a $T=10^5$ K blackbody, we
find that $r_{\rm IF}/R_s$ = 1.08, which is representative of the
codes in the upper range of RT06.

@@

The growth of the ionization front radius is shown in Fig.
\ref{fig:test5_1}, using both $T = 10^4$ K and $x_e = 0.5$ as
ionization front definitions, compared to the analytical solution
(Eq. \ref{eqn:test5_final}).  We define this alternative measure
because the ionization front becomes broad as the D-type front creates
a shock.  Densities in this shock, as seen in Fig.
\ref{fig:test5_2}, are high enough for the gas to recombine but not
radiatively cool.  Before $2t_{\rm rec} \approx 250$ Myr, the
temperature cutoff overestimates $r_s$ by over 10\%; however at later
times, it provides a good match to the $t^{4/7}$ growth at late times.
With the $x_e = 0.5$ criterion for the ionization front, the radius is
always underestimated by $\sim20\%$.  This behavior was also seen in
RT09.
